Modelling the emergence of whisker barrels

### 2019-12-21 paper/briefpaper/brief.tex
Minor whitespace wrangling
Squeezed some more blood out

% It's about how a somatotopically ordered pattern can be
% re-established in the cortex based on information carried with
% thalamocortical axons afferent from the ventrobasal thalamus, but
% without specifying absolute positions as centre-points for the
% barrels. Everything in the paper should be tailored towards arguing
% this idea.

@@

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Sebastian~S.~James}
\author[b]{Leah A. Krubitzer}
\author[a]{Stuart~P.~Wilson}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the constraints under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps the most distinctive pattern can be found in the barrel cortex of many rodent species \cite{woolsey_structural_1970}. The barrels are identifiable soon after birth in layer 4 of primary somatosensory cortex as dense clusters of thalamocortical axons, which are enclosed by borders a few neurons thick from postnatal day 3 \citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the barrels constitute a somatotopic map of the whiskers, with cells within adjacent barrels responding most strongly and quickly to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel patterning reflects subcortical whisker maps comprising cell aggregates called barrelettes in the brainstem and barreloids in the thalamus.
%\cite{van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex network of axon guidance molecules such as ephrin-A5 and A7 and adhesion molecules such as cadherin-6 and 8 \citep{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}.
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17 and transcription factors Emx2, Pax6, Sp8, and Coup-tf1 \citep{shimogori_fibroblast_2005}, which are expressed in gradients that mark orthogonal axes and can be manipulated to stretch, shrink, shift, and even duplicate barrels \cite{assimacopoulos_fibroblast_2012}.
% Removed ref: ,bishop_regulation_2000 after ..Coup-tf1

The barrel boundaries form a Voronoi tessellation across the cortical sheet \citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is preserved in the projection of thalamocortical axons into the cortex, and that a barrel forms by lateral axon branching from an initial centre-point that ceases upon contact with axons branching from adjacent centres. But the assumption of pre-arranged centres is difficult to resolve with the observation that axons arrive in the cortical plate as an undifferentiated bundle, prior to barreloid formation \cite{agmon_organized_1993}.

Alternatively, reaction-diffusion dynamics could generate a Voronoi tessellation without pre-arranged centres, by amplifying characteristic modes in a noisy initial distribution of axon branches, as a net effect of short-range cooperative and longer-range competitive interactions. Accordingly, the barrel pattern would be determined by the relative strength of these interactions and by the shape of the cortical field boundary. However, intrinsic cortical dynamics alone cannot account for the topographic correspondence between thalamic and cortical domains, the irregular sizes and specific arrangement of the barrels in rows and arcs, or the influence of gene expression gradients.

The centre-point and reaction-diffusion models are not mutually exclusive. Pre-organised centres could bias reaction-diffusion processes to generate specific  arrangements more reliably, and mechanisms of lateral axon branching may constitute the tension between cooperation and competition required for self-organization. However, proof that barrel patterning can emerge from an undifferentiated bundle of  axons, based only on local interactions, would show that a separate stage and/or extrinsic mechanism for pre-organising thalamocortical connections need not be assumed. To this end, we ask whether barrel maps can emerge in a system with reaction-diffusion dynamics, under the guidance of signalling gradients, and in the absence of pre-defined centres.

@@

Software for modelling the cortical sheet as a hexagonal lattice with no-flux
boundary conditions and an arbitrary boundary shape, for computing the flux
using a finite difference method, for numerical integration by the
fourth-order Runge-Kutta method, and for reproducing these results is available at \url{https://github.com/ABRG-Models/BarrelEmerge}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\textwidth]{./MainFig.pdf}
\end{center}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1 by \cite{zheng_signal_2001}, with black lines to delineate barrels and to measure departure (Honda-$\delta$; see \cite{senft_mouse_1991}) from a perfect Voronoi tesselation. Right shows the initial distribution of axon branching density ($a$) for one thalamocortical projection, and two molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41 projections by the lengths of green and blue arrows respectively, assuming that similar fields aligned to the posterior-anterior and medial-lateral axes in the ventroposterior medial nucleus of the thalamus are sampled at the locations of putative barreloid centers (reconstructed from \cite{haidarliu_size_2001}). \textbf{C} Simulation results for parameters $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$, $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic projection for which the connection density is maximal, black lines delineate boundaries, and overlaid contours show $c>0.5$ (see Movie S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from simulation approaching that obtained from barrels in \textbf{A} (dotted line);  black squares measure the correspondence between the real and simulated barrel shapes, $\eta$ (the product of the sum of squared differences between real and simulated centers and the sum of differences in area; units mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern in a Fgf8 misexpression experiment (c.f. \cite{assimacopoulos_fibroblast_2012}), simulated by reflecting $\rho_1$ at the join ($\epsilon=80$). All scale bars 1mm.}
\label{fig:main}
\end{figure*}

First we verified that all results established by \cite{karbowski_model_2004} for a 1D axis could be reproduced using our extension to a 2D cortical sheet. Using an elliptical boundary with $M=3$ offset guidance gradients aligned to the longer axis, $N=5$ thalamocortical projections gave rise to five distinct cortical fields at locations that preserved the topographic ordering defined by the original $\gamma$ values. However, we found that specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling fields. As the number of distinct guidance fields is unlikely to approach the number of barrels, further extensions were required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between thalamocortical projections for a limited availability of cortical connections. To introduce competition also in terms of axon branching we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other projections are dense. Divisive normalization keeps the axon branch density bounded on each iteration to the initial density  at $t=0$;
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma},
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$.

The only differences between thalamic projections are their strengths of interaction with the guidance fields, $\gamma$, hence any reliable differences between emergent cortical fields must be due to differences in these values only. We speculate that a projection interacts with a given ephrin field in the cortical subplate with a strength determined by the concentration of a similar molecule at its thalamic origin, i.e., the putative barreloid center. As such, two orthogonal linear thalamic gradients were defined, from which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41 barreloid centers recreated from \cite{haidarliu_size_2001} (Fig.\,1B).

A cortical boundary enclosing 41 corresponding barrels was traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}, and Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the resulting domain, using $M=2$ linear signalling gradients aligned with the anterior-posterior and medial-lateral axes. From random initial conditions for $a({\bf x},0)$ and $c({\bf x},0)$, a clear Voronoi-like tesselation emerged (Fig.\,1C; see Movie S1). A reduction in the Honda delta metric (see \cite{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged within $\approx 3000$ iterations (see Fig.\,1D). A measure of the difference in shape between real and simulated barrels revealed a strong correspondence (see Fig.\,1D).

To further investigate the interplay of genetic and intrinsic mechanisms we simulated a seminal barrel duplication experiment \cite{shimogori_fibroblast_2005},  first by solving on a cortical domain comprising two separate, identically shaped cortical boundaries. In this case one distinct cortical field per thalamic projection emerged, within one of the two boundaries only, i.e., no duplication. However, merging the two fields to create an extended mirror-symmetric boundary shape, with an anterior-posterior guidance field reflected at the join, gave rise to two mirror-symmetrical barrel fields comprising 2N barrels, i.e., two cortical fields for each thalamic projection (Fig.\,1E). Together these results suggest that by the misexpression of Fgf8, Shimogori \& Grove \citep{shimogori_fibroblast_2005} effectively created one large barrel field rather than two distinct cortical fields.

@@

The present results suggest that the key requirements for the emergence of realistic barrel patterning are i) at each cortical location thalamocortical projections compete for a limited number of available synaptic connections (Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a given projection is reduced by the density of other projections (Eq.~\ref{eq:comp}), and iii) the branch density of each projection is conserved over time (Eq.~\ref{eq:norm}).

The emergence of barrels in simulation required competition between thalamic projections in terms of synaptic connectivity, and also in terms of cortical space (as represented by $\chi$). This requirement may account for the absence of barrels in rodents with larger brains, such as capybara, for which competition for space is presumably weaker \citep{Woolsey1975}. Hence, irrespective of whether barrels are necessary for adaptive whisker function, the emergence of somatotopically ordered modular structures may be an inevitable consequence of local competition for cortical territory driven by input from an array of discrete sensory organs \citep{purves_iterated_1992}.

It is important to emphasize that the formulation of the model is entirely local, insofar as simulation requires no information to be communicated from a given cortical grid cell to any but those immediately adjacent (via diffusion). Hence the simulations demonstrate how a self-organising system, constrained by genetically specified guidance cues and by the shape of the cortical field boundary, can faithfully reproduce an arrangement of cell aggregates in one neural structure as a topographic map in another.

Moreover, the present results confirm that somatotopic map formation does not require the pre-specification of centre-points by as yet undetermined additional developmental mechanisms.

>>>>

% It's about how a somatotopically ordered pattern can be
% re-established in the cortex based on information carried with
% thalamocortical axons afferent from the ventrobasal thalamus, but
% without specifying absolute positions as center-points for the
% barrels. Everything in the paper should be tailored towards arguing
% this idea.

@@

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Sebastian~S.~James}
\author[b]{Leah~A.~Krubitzer}
\author[a]{Stuart~P.~Wilson}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the
constraints under which brains evolve and develop
\citep{purves_iterated_1992}. A distinctive example is found in the barrel
cortex of many rodent species. The barrels
become identifiable in layer 4 of primary somatosensory cortex as clusters of
thalamocortical axons, enclosed by borders a few neurons thick from postnatal
day 3 \citep{erzurumlu_development_2012}. Across the
cortical surface the barrels constitute a somatotopic map of the whiskers,
with cells within adjacent barrels responding most strongly and quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel
patterning reflects subcortical whisker maps comprising cell aggregates called
barrelettes in the brainstem and barreloids in the thalamus.
%\cite{van_der_loos_barreloids_1976}.

Barrel formation requires afferent input from whisker stimulation and thalamic calcium waves \citep{anton-bolanos_prenatal_2019}, and depends on a complex network of axon guidance molecules such as ephrin-A5 and A7 and adhesion molecules such as cadherin-6 and 8 \citep{miller_epha7-ephrin-a5_2006}.
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17 and transcription factors Emx2, Pax6, Sp8, and Coup-tf1 \citep{shimogori_fibroblast_2005}, which are expressed in gradients that mark orthogonal axes and can be manipulated to stretch, shrink, shift, and even duplicate barrels \cite{assimacopoulos_fibroblast_2012}.
% Removed ref: ,bishop_regulation_2000 after ..Coup-tf1

The barrel boundaries form a Voronoi tessellation across the cortical sheet \citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is preserved in the projection of thalamocortical axons into the cortex, and that a barrel forms by lateral axon branching from an initial center-point that ceases upon contact with axons branching from adjacent centers. But the assumption of pre-arranged centers is difficult to resolve with the observation that axons arrive in the cortical plate as an undifferentiated bundle, prior to barreloid formation \cite{agmon_organized_1993}.

Alternatively, reaction-diffusion dynamics could generate a Voronoi tessellation without pre-arranged centers, by amplifying characteristic modes in a noisy initial distribution of axon branches, as a net effect of short-range cooperative and longer-range competitive interactions. Accordingly, the barrel pattern would be determined by the relative strength of these interactions and by the shape of the cortical field boundary. However, intrinsic cortical dynamics alone cannot account for the topographic correspondence between thalamic and cortical domains, the irregular sizes and specific arrangement of the barrels, or the influence of gene expression gradients.

The center-point and reaction-diffusion models are not mutually exclusive. Pre-organised centers could bias reaction-diffusion processes to generate specific  arrangements more reliably, and mechanisms of lateral axon branching may constitute the tension between cooperation and competition required for self-organization. However, proof that barrel patterning can emerge from an undifferentiated bundle of  axons, based only on local interactions, would show that a separate mechanism for pre-organising thalamocortical connections need not be assumed. To this end, we ask whether barrel maps can emerge in a system with reaction-diffusion dynamics, under the guidance of signalling gradients, and in the absence of pre-defined centers.

@@

%All software required to reproduce these results is available at \url{https://github.com/ABRG-Models/BarrelEmerge}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=17.8cm]{./MainFig.png}
\end{center}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \cite{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \cite{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). Software for modelling the cortical sheet as a hexagonal lattice with
  no-flux boundary conditions and an arbitrary boundary shape, for computing
  the flux using a finite difference method, for numerical integration by the
  fourth-order Runge-Kutta method, and for reproducing these results is
  available at \url{https://github.com/ABRG-Models/BarrelEmerge} \textbf{D} Red
  dots show the Honda-$\delta$ metric obtained from simulation approaching
  that obtained from barrels in \textbf{A} (dotted line); black squares
  measure the correspondence between the real and simulated barrel shapes,
  $\eta$ (the product of the sum of squared differences between real and
  simulated centers and the sum of differences in area; units
  mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern in a Fgf8
  misexpression experiment (c.f. \cite{assimacopoulos_fibroblast_2012}),
  simulated by reflecting $\rho_1$ at the join ($\epsilon=80$). All scale bars
  1mm.}
\label{fig:main}
\end{figure*}

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced in our 2D implementation. Using an elliptical boundary with $M=3$ offset guidance gradients aligned to the longer axis, $N=5$ projections gave rise to five distinct cortical fields at locations that preserved the topographic ordering defined by the original $\gamma$ values. However, we found that specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling fields. As the number of distinct guidance fields is unlikely to approach the number of barrels, further extensions were required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
projections for limited cortical connections. To introduce axon branching
competition we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other projections are dense. Divisive normalization keeps the axon branch density bounded on each iteration to the initial density  at $t=0$;
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma},
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$.

The only differences between thalamic projections are their strengths of interaction with the guidance fields, $\gamma$, hence any reliable differences between emergent cortical fields must be due to differences in these values only. We speculate that a projection interacts with a given ephrin field in the cortical subplate with a strength determined by the concentration of a similar molecule at its thalamic origin, i.e., the putative barreloid center. As such, two orthogonal linear thalamic gradients were defined, from which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41 barreloid centers recreated from \cite{haidarliu_size_2001} (Fig.\,1B).
%
A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a({\bf x},0)$, a clear Voronoi-like tesselation emerged (Fig.\,1C; Movie
S1). The Honda $\delta$ metric (see \cite{senft_mouse_1991}) confirmed that a
`good' Voronoi pattern emerged within $\approx 3000$ iterations (Fig.\,1D). A
measure of the difference in shape between real and simulated barrels revealed
a strong correspondence (see Fig.\,1D).

To further investigate the interplay of genetic and intrinsic mechanisms we
simulated a seminal barrel duplication experiment
\cite{shimogori_fibroblast_2005}, by solving on a domain comprising two
separate, identically shaped cortical boundaries. In this case one distinct
cortical field per thalamic projection emerged, within one of the two
boundaries only, i.e., no duplication. However, merging the two fields to
create an extended mirror-symmetric boundary shape, with an anterior-posterior
guidance field reflected at the join, gave rise to two mirror-symmetrical
barrel fields comprising $\approx 2N$ barrels, i.e., two cortical fields for
each projection (Fig.\,1E). Together these results suggest that by the
misexpression of Fgf8, Shimogori \& Grove \citep{shimogori_fibroblast_2005}
effectively created one large, rather than two distinct, cortical fields.

% Omit to shorten manuscript
%\subsection*{Movie S1 caption}
%
%Movie corresponding to Fig.\,1C in the main paper. Simulation parameters were
%$N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
%$\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
%projection for which the connection density is maximal, black lines delineate
%boundaries, and overlaid contours show $c>0.5$. The final frame in the movie
%is step 25000 of the simulation.

@@

These results suggest that the emergence of realistic barrel patterning
requires that i) at each cortical location thalamocortical projections compete for a limited number of available synaptic connections (Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a given projection is reduced by the density of other projections (Eq.~\ref{eq:comp}), and iii) the branch density of each projection is conserved over time (Eq.~\ref{eq:norm}).

The emergence of barrels in simulation required competition between projections for synaptic connectivity, and also for cortical space (represented by $\chi$). This requirement may account for the absence of barrels in rodents with larger brains, such as capybara, for which competition for space is presumably weaker \citep{Woolsey1975}. Hence, irrespective of whether barrels are necessary for adaptive whisker function, the emergence of somatotopically ordered modular structures may be an inevitable consequence of local competition for cortical territory driven by input from an array of discrete sensory organs \citep{purves_iterated_1992}.

We emphasize that the formulation of the model is entirely local; the
simulation permits information to be communicated from a given cortical grid
cell only to those immediately adjacent (via diffusion). Hence the simulations
demonstrate how a self-organising system, constrained by genetically specified
guidance cues and by the shape of the boundary, can faithfully reproduce an
arrangement of cell aggregates in one neural structure as a topographic map in
another.

Moreover, the present results confirm that somatotopic map formation does not require the pre-specification of center-points by as yet undetermined additional developmental mechanisms.

\acknow{The authors thank Jason Berwick for advice and data used to construct
  Fig.\,1A. This work was supported by the James S.~McDonnell Foundation
  (grant 220020516).}

\showacknow{} % Display the acknowledgments section
